Astr 513: Final Project

### 2016-05-01 Final_Project/ms.tex
Wrote things for Sections 3 and 4

\usepackage{todonotes}
\newcommand\TM[1]{\todo[color=green!40, inline, size=\small]{TM: #1}}
\newcommand\EW[1]{\todo[color=blue!40, inline, size=\small]{TM: #1}}
\newcommand\LC[1]{\todo[color=magenta!40, inline, size=\small]{TM: #1}}

@@

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
\TM{TODO: Complete this section.}
\EW{You guys might want to read this over and tell me if you think I should
  change/add anything.}
A type \rom{1}a supernova (\sna{}) is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy
\citep[see, e.g.,][chap. 18.4]{ryden2010}, which is enough to overcome the
gravitational energy and the star explodes. Because all \snea{} are caused by a
similar reaction, they all have similar luminosities, and more importantly, the
light curves all have the same shape: a sharp maximum followed by a smooth,
gradual decay. Studying nearby \snea{} has shown that the rate of decline of
the light curve is correlated to the peak luminosity, and this correlation has
been calibrated using Cephied variable stars located in the same galaxies as
nearby \sna{}. With this information, we can estimate peak luminosity of an
\sna{} from its light curve, thereby providing us with a standard candle with
which to measure distance. They were first used in this way by the
Cal\'{a}n/Tololo Supernova Survey, and it was using \snea{} that
\citet{riess1998} and \citet{perlmutter1999} were able to discover the
accelerating expansion rate of the Universe for which three of the authors
later won the Nobel Prize in physics.

\section{Planck 2015 Results}
\label{sec:planck}
In 2009 the European Space Agency launched the \Planck satellite in order
to measure the Cosmis Microwave Background (CMB) with unparalleled precision.
By measuring the small anisotropies in the temperature of the CMB across
the entire sky one calculates the angular power spectrum, which can be
modeled using first principles (see, e.g., \citet{dodelson2003}).
By using a likelihood method similar to our own, the authors of
\citet{planck2013} estimated 14 nuisance parameters and 
six cosmological parameters, including the geometrical parameters,
$\Omega_M$, $\Omega_\Lambda$ and $\Omega_k$, that we use in our
SNIa analysis. The only prior used by \citet{planck2013}
on cosmological parameters was from the WMAP result \citep{wmap2011}
for the optical depth $\tau$. We do not use this parameter
in our SNIa model.

@@

\section{Data}
\label{sec:data}
\TM{TODO: Complete this section.}
Our data is compiled from \citet{betoule2014} and \citet{rest2014} blah blah blah...

@@

\citet{betoule2014} define their ``joint light-curve analysis'' (JLA) 
sample, a total of 740 \snea{} compiled from a combination of data from the Sloan Digital Sky 
Survey (SDSS), the Supernova Legacy Survey (SNLS), the Hubble Space 
Telescope (HST), and several other experiments. The 613 SDSS \snea{} 
come from the SDSS supernova survey (SDSS-SNS) of SDSS-II \citep{sako2014} 
with redshifts of $0.05 < z < 0.4$, and 374 of them are selected from those 
that were confirmed in the spectroscopic follow-up. The rest of the 
\sne{} come from the compilation by \citet{conley2011}, which includes 
242 spectroscopically confirmed \snea{} from SNLS at redshifts $0.2 < z < 1$, 
14 \snea{} from HST at very high redshifts of $0.7 < z < 1.4$, and 123 
low redshift ($z < 0.08$) from a variety of sources. We made use of the 
publicly available distance moduli, redshifts, and covariance 
matrix provided by \citet{betoule2014}\footnote{Data is available 
for download via \url{http://cdsweb.u-strasbrg.fr/cgi-bin/qcat?J/A+A/568/A22}}. 
We note that the distance moduli are binned, so that they are only computed and 
given on a grid of redshifts. The distance modulus can then be well 
approximated by a piecewise linear function of $\log(z)$ that is given 
in a single redshift bin $z_b \le z < z_{b+1}$ as 
\citep[see][Appendix E.1]{betoule2014}
\begin{equation}
\label{eq:binnedMu}
\bar{\mu}(z) = (1 - \alpha) \mu_b + \alpha \mu_{b+1},
\end{equation}
where they have defined 
$\alpha = \log\left(\frac{z}{z_b}\right)/\log\left(\frac{z_{b+1}}{z_b}\right)$ 
and $\mu_i$ is the distance modulus at control point $z_i$. 
Therefore, the data and covariance matrix are merely given at the 
31 control points log-spaced in the redshift range $0.01 < z < 1.3$.

\subsection{Covariances}
\label{sec:covariances}
\TM{A subsection that explains how the covariance matrix looks. Basically
  a brief discussion about how it was made.}

@@

@@

\section{Distance Modulus Model}
\label{sec:model}
\TM{This explains how to go from a cosmology to $\mu(z)$.}

>>>>

\usepackage{todonotes}
\newcommand\TM[1]{\todo[color=green!40, inline, size=\small]{TM: #1}}
\newcommand\EW[1]{\todo[color=blue!40, inline, size=\small]{EW: #1}}
\newcommand\LC[1]{\todo[color=magenta!40, inline, size=\small]{LC: #1}}

@@

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
\EW{You guys might want to read this over and tell me if you think I should
  change/add anything.}
A type \rom{1}a supernova (\sna{}) is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy
\citep[see, e.g.,][chap. 18.4]{ryden2010}, which is enough to overcome the
gravitational energy and the star explodes. Because all \snea{} are caused by a
similar reaction, they all have similar luminosities, and more importantly, the
light curves all have the same shape: a sharp maximum followed by a smooth,
gradual decay. Studying nearby \snea{} has shown that the rate of decline of
the light curve is correlated to the peak luminosity, and this correlation has
been calibrated using Cephied variable stars located in the same galaxies as
nearby \sna{}. With this information, we can estimate peak luminosity of an
\sna{} from its light curve, thereby providing us with a standard candle with
which to measure distance. They were first used in this way by the
Cal\'{a}n/Tololo Supernova Survey, and it was using \snea{} that
\citet{riess1998} and \citet{perlmutter1999} were able to discover the
accelerating expansion rate of the Universe for which three of the authors
later won the Nobel Prize in physics.

\section{Planck 2015 Results}
\label{sec:planck}
In 2009 the European Space Agency launched the \Planck satellite in order
to measure the Cosmis Microwave Background (CMB) with unparalleled precision.
By measuring the small anisotropies in the temperature of the CMB across
the entire sky one calculates the angular power spectrum, which can be
modeled using first principles (see, e.g., \citet{dodelson2003}).
By using a likelihood method similar to our own, the authors of
\citet{planck2013} estimated 14 nuisance parameters and 
six cosmological parameters, including the geometrical parameters,
$\Omega_M$, $\Omega_\Lambda$ and $\Omega_k$, that we use in our
{\sna} analysis. The only prior used by \citet{planck2013}
on cosmological parameters was from the WMAP result \citep{wmap2011}
for the optical depth $\tau$. We do not use this parameter
in our {\sna} model.

@@

\section{Data}
\label{sec:data}
Our {\sna} data is compiled from the samples analyzed in \citet{betoule2014} and \citet{rest2014}. In the following subsections, we will specify the surveys and detail the composition of the two samples. Additionally, we will describe the publicly released data from the two papers.

@@

\citet{betoule2014} define their ``joint light-curve analysis'' (JLA) 
sample, a total of 740 \snea{} compiled from a combination of data from the Sloan Digital Sky 
Survey (SDSS), the Supernova Legacy Survey (SNLS), the Hubble Space 
Telescope (HST), and several other experiments. The 613 SDSS \snea{} 
come from the SDSS supernova survey (SDSS-SNS) of SDSS-II \citep{sako2014} 
with redshifts of $0.05 < z < 0.4$, and 374 of them are selected from those 
that were confirmed in the spectroscopic follow-up. The rest of the 
\sne{} come from the compilation by \citet{conley2011}, which includes 
242 spectroscopically confirmed \snea{} from SNLS at redshifts $0.2 < z < 1$, 
14 \snea{} from HST at very high redshifts of $0.7 < z < 1.4$, and 123 
low redshift ($z < 0.08$) from a variety of sources. We made use of the 
publicly available distance moduli, redshifts, and covariance 
matrix provided by \citet{betoule2014}\footnote{Data is available 
for download via \url{http://cdsarc.u-strasbg.fr/viz-bin/qcat?J/A+A/568/A22}}. 
We note that the distance moduli are binned, so that they are only computed and 
given on a grid of redshifts. The distance modulus can then be well 
approximated by a piecewise linear function of $\log(z)$ that is given 
in a single redshift bin $z_b \le z < z_{b+1}$ as 
\citep[see][Appendix E.1]{betoule2014}
\begin{equation}
\label{eq:binnedMu}
\bar{\mu}(z) = (1 - \alpha) \mu_b + \alpha \mu_{b+1},
\end{equation}
where they have defined 
$\alpha = \log\left(\frac{z}{z_b}\right)/\log\left(\frac{z_{b+1}}{z_b}\right)$ 
and $\mu_i$ is the distance modulus at control point $z_i$. 
Therefore, the data and covariance matrix are merely given at the 
31 control points log-spaced in the redshift range $0.01 < z < 1.3$.

\subsection{The Pan-STARRS1 sample}
\label{sec:rest}

\citet{rest2014} uses the first 1.5 year of the first telescope of the Panoramic Survey Telescope and Rapid Response System (Pan-STARRS1) and a compilation of low-redshift {\sna} studies. The low-redshift sample consists of {\snea} collected by Calan/Tololo (Hamuy et al. 1996a), CfA1 (Riess et al. 1999), CfA2 (Jha et al. 2006), CfA3 (Hicken et al. 2009a), CSP (Contreras et al. 2010), and CfA4 (Hicken et al. 2012). After three quality cuts, 113 spectroscopically confirmed {\snea} from Pan-STARRS1 ($0.03 < z < 0.65$) and 222 light curves from 197 low-redshift {\snea} are used for cosmological analysis.

The publicly available data from \citet{rest2014} contains all light curves, light curve fit information, covariance matrix, and each {\sn} host information.\footnote{Data can be accessed at \url{http://ipp0012.ifa.hawaii.edu/SN1aLC/}}. For our analysis, we utilized the redshifts, distance moduli, and covariance matrix of the final sample of 310 {{\snea} or 335 light curves.

\subsection{Covariances}
\label{sec:covariances}
\TM{A subsection that explains how the covariance matrix looks. Basically
  a brief discussion about how it was made.}
\LC{Wrote about Rest's cov matrix. Still need description of Betoule's.}

@@

The publicly released covariance matrix from \citet{rest2014} is comprised of systematic errors for the final sample. These systematics, defined in \citet{scolnic2014}, include flux calibration errors, selection effects, light curve fitting, host galaxy relations, Milky Way extinction corrections, and coherent velocity flows. Calculating the covariance matrix involves varying a systematic parameter and determining the difference between the original distance and the new distance. The covariance between {\sn} $i$ and {\sn} $j$ is 
\begin{equation}
\label{eq:RestCovMatrix}
C_{ij}=\displaystyle \sum_{k = 1}^K \left(\frac{\partial \mu_i}{\partial S_k}\right)\left(\frac{\partial \mu_j}{\partial S_k}\right)\left(\sigma_{S_k}\right)^2,
\end{equation}
where the summation is over $K$ systematics, $\sigma_{S_k}$ is each systematic error magnitude, and $\partial \mu$ is the difference in distance modulus from varying a systematic parameter.

@@

\section{Distance Modulus Model}
\label{sec:model}
\TM{This explains how to go from a cosmology to $\mu(z)$.}
\LC{Need more here but I wrote most of the relevant eqns}
The main objective of \citet{betoule2014} and \citet{rest2014} is to place constraints on cosmological parameters $\Omega_M$, $\Omega_\Lambda$, and $w$. In order to so, we must examine the relationship between distance modulus and cosmology. The distance modulus $\mu$ is determined from the luminosity distance $d_L$ by 
\begin{equation}
\label{eq:mu_dL}
\mu=5 \text{log}(d_L)+25. 
\end{equation}
The luminosity distance depends on the cosmology by 
\begin{equation}
\label{eq:luminosityDistance}
d_L(z; w, \Omega_M, \Omega_\Lambda, H_0)= (1+z)|\Omega_k|^{-1/2}\mathcal{S}_{k}\displaystyle\left[c|\Omega_k|^{1/2}\int_{0}^{z}\frac{dz'}{H(z')}\right],
\end{equation}
where $\Omega_k \equiv 1-\Omega_M-\Omega_{\Lambda}$. The function $\mathcal{S}_{k}$ depends on the curvature density $\Omega_k$ in that 
\[
    \mathcal{S}_{k}= 
\begin{cases}
    \text{sin}(x),&  \Omega_k < 0\\
    x,& \Omega_k = 0\\
    \text{sinh}(x),& \Omega_k > 0
\end{cases}
\]
and the function $H(z)$ is 
\begin{equation}
\label{eq:hubbleFunction}
H(z)= H_0\displaystyle\left[\Omega_M(1+z)^3+\Omega_{\Lambda}(1+z)^{3(1+w)}+\Omega_k(1+z)^2\right]^{1/2}.
\end{equation}
Three (?) cosmologies are of interest to us: \LC{Not sure which ones we're interested in. $\Lambda CDM$, $\Omega_M=1$, ?}
